---
title: "106P Final Project"
author: "Jack Brun, Michael Dalton"
date: "2025-05-17"
output:
  pdf_document:
    latex_engine: xelatex
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 2. Merger analysis 1: HHI Indices

### Start by describing how concentrated markets (defined as routes) are. On how many of the routes is both Alaskan and Hawaiian present in 2024? Compute the pre-merger HHI (Herfindahl index) on each route in 2024. What is the average and median HHI prior to the merger? Next, assume Alaskan and Hawaiian merge in 2024. How high is the post-merger HHI? Compute the change in the HHI for every market (route). Based on the 2010 Merger Guidelines, how many markets are green, yellow, red? (see topic 5 slide 16). What does this imply for the likelihood of further investigation into the merger?

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#clearall
rm(list=ls()) 
```

```{r}
setwd("/Users/jackbrun/Desktop")
getwd() 
```

```{r}
# Load data
airlinef <-read.csv("airline_data_firmlevel.csv", sep =  "," , header=TRUE)
airline <-read.csv("airline_data_marketlevel.csv", sep = "," , header=TRUE)
routes <-read.csv("db1bmarket.csv", sep = "," , header=TRUE)
```

```{r question 2}
# How many routes is both Alaskan and Hawaiian present in 2024
# Filter for Alaska and Hawaiian Airlines routes
alka_hai_routes <- subset(routes, rpcarrier %in% c("AS", "HA"))

# Count how many unique carriers are present per route
route_carriers <- aggregate(rpcarrier ~ mkt, data = alka_hai_routes, FUN = function(x) length(unique(x)))

# Keep only routes where both AS and HA fly
shared_routes <- subset(route_carriers, rpcarrier == 2)

# Show results
number_shared_routes <- nrow(shared_routes)
print(paste("Number of routes with both Alaska & Hawaiian airlines:", number_shared_routes))
```

The Herfindahl-Hirschman Index (HHI) is calculated using the formula:
$$
HHI = 10{,}000 \sum_j s_j^2
$$
where \( s_j \) is the market share of firm \( j \) in a given market.

```{r}
# Calculate pre-merger HHI for each route
hhi_per_route <- aggregate(ms ~ mkt, data = routes, FUN = function(x) 10000 * sum(x^2))

# Define the result column to HHI & view average and median HHI
colnames(hhi_per_route)[2] <- "HHI"
average_hhi <- mean(hhi_per_route$HHI)
median_hhi <- median(hhi_per_route$HHI)

# Print
print(paste("Average pre-merger HHI:", round(average_hhi, 2)))
print(paste("Median pre-merger HHI:", round(median_hhi, 2)))
```
Hawaiian and Alaska potential merger in 2024:
```{r}
# Original data
df_post <- routes

# Merge data Alaska and Hawaiian together & merge market shares post merger
df_post$rpcarrier_merged <- ifelse(df_post$rpcarrier %in% c("AS", "HA"), "ASHA", df_post$rpcarrier)
merged_df <- aggregate(ms ~ mkt + rpcarrier_merged, data = df_post, FUN = sum)

# Post merger HHI
hhi_post <- aggregate(ms ~ mkt, data = merged_df, FUN = function(x) 10000 * sum(x^2))
colnames(hhi_post)[2] <- "HHI_post"

# Merge with pre-merger HHI
hhi_pre <- aggregate(ms ~ mkt, data = routes, FUN = function(x) 10000 * sum(x^2))
colnames(hhi_pre)[2] <- "HHI_pre"

# Merge the two HHI's & calculate HHI change
hhi_combined <- merge(hhi_pre, hhi_post, by = "mkt")
hhi_combined$HHI_change <- hhi_combined$HHI_post - hhi_combined$HHI_pre

# Summary mean and median change in HHI post merger
mean_change <- mean(hhi_combined$HHI_change)
median_change <- median(hhi_combined$HHI_change)

# Summary for post-merger HHI
average_post_hhi <- mean(hhi_post$HHI_post)
median_post_hhi <- median(hhi_post$HHI_post)

# Results
print(paste("Average post-merger HHI:", round(average_post_hhi, 2)))
print(paste("Median post-merger HHI:", round(median_post_hhi, 2)))

# Results
print(paste("Average change in HHI:", round(mean_change, 2)))
print(paste("Median change in HHI:", round(median_change, 2)))

# Define each market based on the DOJ guidelines
hhi_combined$category <- with(hhi_combined, ifelse(HHI_pre < 1500, "Green",
                                         ifelse(HHI_pre < 2500 & HHI_change < 100, "Green",
                                         ifelse(HHI_pre < 2500 & HHI_change >= 100, "Yellow",
                                         ifelse(HHI_pre >= 2500 & HHI_change < 100, "Yellow",
                                         ifelse(HHI_pre >= 2500 & HHI_change >= 200, "Red", "Yellow"))))))

# Sort routes by green, yellow, and red
table(hhi_combined$category)
```
Based on our analysis of the markets jointly served by Alaska Airlines and Hawaiian Airlines, we identified 299 overlapping routes in 2024. The pre-merger mean HHI is 5,729.12 and the median is 5,000.00 which, based on the DOJ guidelines, indicates a very concentrated market with potential monopoly or oligopoly. To assess potential antitrust implications, we calculated the post-merger mean and median HHI, which were 5,748.76 and 5,011.34, respectively. These results are in line with our change in mean and median HHI which were 19.6 and 0 respectively.  These results fall under the “Small Change in Concentration” section defined by the DOJ to be “Mergers involving an increase in the HHI of less than 100 points are unlikely to have adverse competitive effects and ordinarily require no further analysis.” 

In addition, we categorized the total routes into the DOJ's three HHI concentration tiers: Green (HHI < 1,500): 587 routes, Yellow (HHI 1,500–2,500): 5,568 routes, Red (HHI > 2,500): 62 routes. These results reinforce the conclusion that the merger would not significantly alter the competitive landscape. Despite the high pre-merger concentration, the minimal increase in HHI and the limited number of routes in the red zone suggest that the transaction is unlikely to raise substantial antitrust concerns. As outlined in the 2010 Merger Guidelines, no further competitive analysis appears needed.

